---
title: "002 - Random Graphs"
output: html_notebook
---

Creating a randomized graph is actually quite useful. Seeing these graphs with specified numbers of nodes and edges will allow you to quickly get a sense of how connected graphs can be at different sizes.

## Setup

Ensure that the developement version of **DiagrammeR** is installed. Load in the package with `library()` and also ensure that the **magrittr** package is available.

```{r, results=FALSE}
devtools::install_github("rich-iannone/DiagrammeR")

library(DiagrammeR)
library(magrittr)
```

## Part 1. Random Graphs with `create_random_graph()`

The `create_random_graph()` function is provided in **DiagrammeR** with several options for creating random graphs. In all of the examples, the function will pipe the graph object to the `render_graph()` function (using the **magrittr** `%>%` operator) with the `output = "visNetwork"` option set. This is useful for quickly inspecting the graph upon creation.

To start, we can create a not-so-random graph with 2 nodes (`n` argument) and 1 edge (`m` argument). By default, the graph produced is an undirected graph.

```{r}
create_random_graph(n = 2, m = 1) %>% render_graph
```

It's obviously better to make a graph with more nodes and edges. Next, we will create a random graph with 15 nodes and 30 edges:

```{r}
create_random_graph(n = 15, m = 30) %>% render_graph
```

In all of the random graphs created with this function, there is only one edge created for each pair of nodes (i.e., no cases with multiple edges between nodes).

If you specify a number of edges that exceeds the number in a fully-connected graph of size `n`, you get an error. It's an informative error (providing the maximum number of edges `m` for the given `n`) but it's an error nonetheless.

```{r}
create_random_graph(n = 15, m = 200) %>% render_graph
```

So, using `n = 15` and `m = 105` will yield a fully-connected graph with 15 nodes:

```{r}
create_random_graph(n = 15, m = 105) %>% render_graph
```

Alternatively, you don't need to have edges in the randomly generated graph. Simply specify `m = 0` for any number of nodes `n`:

```{r}
create_random_graph(n = 512, m = 0) %>% render_graph
```

# Part 2. Random Yet Reproducible

Setting a seed is a great way to create something random yet reproduces that random something later use. This can be done within the `create_random_graph()` function by specifying a seed number with the argument `set_seed`. Here's an example

```{r}
create_random_graph(n = 4, m = 4, set_seed = 30) %>% render_graph
```

Upon repeat runs, the connections in the graph will be the same each and every time (e.g., node `1` is attached to all other nodes, `2` is connected to `1`, etc.).

## Part 3. Directed/Undirected Randomness

By default, the random graphs generated are undirected. To produce directed graphs, simply include `directed = TRUE` in the `create_random_graph()` statement.

```{r}
create_random_graph(n = 15, m = 22, directed = TRUE) %>% render_graph
```
