---
title: "003 - The Node Type and the Edge Relationship"
output: html_notebook
---

Sometimes, you want some collection of nodes in your graph to be part of a group. In such a case, we specify a node attribute called `type`. Edges can be provided with a grouping name as well. This is with the `rel` edge attribute (which is short for 'relationship').

## Setup

Ensure that the developement version of **DiagrammeR** is installed. Load in the package with `library()`.

```{r load_packages, results=FALSE}
# devtools::install_github("rich-iannone/DiagrammeR")

library(DiagrammeR)
```

## Part 1. The Node `type` Attribute

When creating a node data frame, we easily apply groupings for nodes by supplying values for the `type` argument. 

```{r create_ndf}
# Create a node data frame with 8 nodes, and
# assign several `type` values and unique
# `label` values
ndf <- 
  create_node_df(
    n = 8,
    type = c("X", "X", "X", "X",
             "Y", "Y", "Z", "Z"),
    label = c("dfs", "udd", "wkd", "xkd",
              "qpx", "mdj", "ldk", "wjq"))

# View the node data frame
ndf
```

```{r render_graph_nodes}
# Create a graph with just nodes (no edges):
graph <-
  create_graph(nodes_df = ndf)

graph %>% render_graph(output = "visNetwork")
```

By assigning `type` values for each of the nodes, they are automatically colored when rendered by `visNetwork` (in this case blue, yellow, and red for types 'X', 'Y', and 'Z') and this is quite convenient.

There is a function that allows you to get a count of all nodes, or, counts of nodes with specific `type` values: the `node_count()` function. To get a count of all nodes, simply pass the graph object to `node_count()`. The graph object can be provided to `node_count()` with a **magrittr** `%>%`.

```{r get_node_count}
graph %>% node_count
```

To get a count of nodes grouped by `type`, use `type = TRUE`. A named vector will be returned (where the names are the `type` values).

```{r get_node_count_type}
graph %>% node_count(type = TRUE)
```

To get a count of nodes of a specific type, use the `type` argument again but this time supply the `type` value(s).

```{r get_node_count_type_x_y}
# Get a count of nodes of type `X`
graph %>% node_count(type = "X")

# Get a total count of nodes of the
# `X` and `Y` types
graph %>% node_count(type = c("X", "Y"))
```

To identify which nodes are of a specific type, use the `get_node_ids()` function with `type` supplied to the `node_attr` argument and also supply `X` to the `match` argument. This returns a vector of node ID values.

```{r get_node_ids}
graph %>% get_node_ids("type == 'X'")
```

You can also return the node IDs for nodes of several `types` by combining `get_node_ids()` statements with `c()`.

```{r get_node_ids_c}
graph %>% 
{
  c(get_node_ids(., "type == 'X'"),
    get_node_ids(., "type == 'Y'"))
}
```

## Part 2. The `edge` Relationship

The analogue to `type` for edges is the `rel` attribute. Here is an example of an edge data frame where relationship values are provided for all edges.

```{r create_edf}
# Create an edge data frame
edf <- 
  create_edge_df(
    from = c(1, 2, 5, 7, 6, 8, 4, 4),
    to =   c(2, 5, 8, 8, 8, 3, 3, 1),
    rel = c("rel_a", "rel_a", "rel_b",
            "rel_c", "rel_b", "rel_a",
            "rel_b", "rel_c"))

# View the edge data frame
edf
```

Create a graph object that contains nodes with `type` attrs and edges with `rel` attrs, and, view the graph in the Viewer.

```{r create_graph_ndf_edf}
# Create the graph using the
# ndf and edf objects
graph <-
  create_graph(
    nodes_df = ndf,
    edges_df = edf)

# Render the graph
graph %>% render_graph(output = "visNetwork")
```


As can be seen, all the edges have `rel` values displayed when using `output = "visNetwork"`.

The edge analogue to `node_count()` is `edge_count()` and this function allows you to get counts of all edges or those edges by relationship. To get a count of all edges, simply pass the graph to `edge_count()`.

```{r get_edge_count}
edge_count(graph)
```

By default, the `rel` argument is set to `FALSE` and that provides a total count of edges. Setting `rel` to `TRUE` yields a named vector of edge counts by relationship.

```{r get_edge_count_rel_true}
edge_count(graph, rel = TRUE)
```

Again, as with `node_count()`, one or more specified `rel` values can be used to determine the number of matching edges.

```{r get_edge_count_with_rel}
# Get a count of edges of with rel `rel_a`
edge_count(graph, rel = c("rel_a"))

# Get a total count of nodes of the
# `rel_a` and `rel_b` edges
edge_count(
  graph, rel = c("rel_a", "rel_b"))
```

To identify which edges are of a particular rel, use the `get_edges()` function with the `rel` supplied. This function allows you have the edges returned in multiple ways by use of the `return_type` argument. Output can either be as a `vector` (the default), as a data frame (`df`), or as a `list`.

```{r get_edges_with_condition_vec}
graph %>%
  get_edges(
    "rel == 'rel_a'")
```

```{r get_edges_with_condition_df}
graph %>%
  get_edges(
    "rel == 'rel_a'",
    return_type = "df")
```

```{r get_edges_with_condition_list}
graph %>%
  get_edges(
    "rel == 'rel_a'",
    return_type = "list")
```

Either way, you can indeed isolate those edges that have certain `rel` values attached.

