---
title: "006 - Graph Traversals"
output: html_notebook
---

Graph traversal functions are available in several general-purpose graph libraries and in graph database technologies. Once their use is understood, they are extremely useful for getting the particular data you need from a property graph (a graph with labeled nodes and edges using the `type` and `rel` attributes, respectively). Forward-pipe operations using the **magrittr** `%>%` operator are indispensible here as they make such traversal operations (and concurrent data gathering) much easier to reason about.

## Setup

Ensure that the development version of **DiagrammeR** is installed. Load in the package with `library()`.

```{r load_packages, results=FALSE}
#devtools::install_github("rich-iannone/DiagrammeR")

library(DiagrammeR)
```

## Part 1. Preparing the Graph

The example graph to be used is a fake dataset with contributors to software projects on a platform not unlike **GitHub**. First, get paths to the CSV files available in the package.

```{r}
# This is a CSV file containing contributors to
# software projects:
contributors_csv <-
  system.file("extdata/contributors.csv",
              package = "DiagrammeR")

# Display the column names in this CSV
colnames(read.csv(contributors_csv,
                  stringsAsFactors = FALSE))
```

```{r}
# This is a CSV file containing information about
# the software projects:

projects_csv <-
  system.file("extdata/projects.csv",
              package = "DiagrammeR")

colnames(read.csv(projects_csv,
                  stringsAsFactors = FALSE))
```

```{r}
# This is a CSV file with information about the
# relationships between the projects and their
# contributors:

projects_and_contributors_csv <-
  system.file("extdata/projects_and_contributors.csv",
              package = "DiagrammeR")

colnames(read.csv(projects_and_contributors_csv,
                  stringsAsFactors = FALSE))
```
Create the property graph by adding the CSV data to a new graph; the `add_nodes_from_csv()` and `add_edges_from_csv()` functions are used to create nodes and edges in the graph.

```{r}
# Create the main graph
graph <-
  create_graph() %>%
  set_graph_name("software_projects") %>%
  add_nodes_from_table(
    system.file(
      "extdata", "contributors.csv",
      package = "DiagrammeR"),
    set_type = "person",
    label_col = "name") %>%
  add_nodes_from_table(
    system.file(
      "extdata", "projects.csv",
      package = "DiagrammeR"),
    set_type = "project",
    label_col = "project") %>%
  add_edges_from_table(
    system.file(
      "extdata", "projects_and_contributors.csv",
      package = "DiagrammeR"),
    from_col = "contributor_name",
    to_col = "project_name",
    ndf_mapping = "label",
    rel_col = "contributor_role")
```

Now that the graph object has been generated, render the graph in the **RStudio** Viewer:
```{r}
render_graph(graph, output = "visNetwork")
```
## Part 2. Querying the Graph to Get Answers

Now that we have a property graph, we can find out bits of information without directly inspecting such information (i.e., not working with the graph's internal node or edge data frames). This will be important when a property graph becomes larger or more complex since manual work these data frames gets difficult and impractical.

We can get the average age of all the contributors. We do this by first selecting all nodes of the `type` that is `person`. Then, for these selected nodes, we cache the node attribute corresponding to `age`. The cache is retained in the graph object. The next steps are to retrieve the cache with `get_cache()` and pass that numeric vector to the `mean()` function.

```{r}
graph %>% 
  select_nodes("type == 'person'") %>%
  cache_node_attrs_ws("age", "numeric") %>%
  get_cache() %>%
  mean()
```
The average age is `33.6` years.

To get the total number of commits to all software projects, follow a similar strategy of selecting all edges with `select_edges()`. Cache all of the values for the edge attribute `commits`. Get the cache as a vector with `get_cache()` and compute the sum of those values.

```{r}
graph %>% 
  select_edges() %>%
  cache_edge_attrs_ws("commits", "numeric") %>%
  get_cache() %>% 
  sum()
```

## Part 3. Using Traversals to Get Answers

The traversal functions in **DiagrammeR** are all named as `trav_...()` and their naming indicates the type of traversal. In essence, a traversal will modify a graph selection by taking into account the current selection (either of nodes or edges), performing a traversal with the selection, and creating a new selection of nodes or edges for the successful traversals. Here is a summary that explains what each of the traversal functions does.

| traversal function | type of traversal | description |
|--------------------|-------------------|-------------|
| `trav_out()`       | node to node      | move from a node across any outbound edge(s) to node(s) |
| `trav_in()`        | node to node      | move from a node across any inbound edge(s) to node(s) |
| `trav_both()`      | node to node      | move from a node across any inbound or outbound edge(s) to node(s) |
| `trav_out_edge()`  | node to edge      | move from a node to outbound edge(s)
| `trav_in_edge()`   | node to edge      | move from a node to inbound edge(s)
| `trav_out_node()`  | edge to node      | move from an edge to the node where the edge is outbound
| `trav_in_node()`   | edge to node      | move from an edge to the node where the edge is inbound

Now it is time to use traversals to get more specific information from the graph. If we were to get the total number of commits from Josh, we can first select the node for Josh using `select_nodes()` and specifying that the `name` attribute should have the value `Josh`. Commits are values that exist on the graph's edges, so, we must traverse from the single node (that is, the current node selection) onto any outbound edges. This is done with the `trav_out_edge()` function, and using it will modify the selection to now contain any of those edges directed out of the node.

```{r}
graph %>% 
  select_nodes("name == 'Josh'") %>%
  trav_out_edge() %>%
  cache_edge_attrs_ws("commits", "numeric") %>%
  get_cache() %>% 
  sum()
```

Switching to another user in the graph, get the total number of commits from Louisa:

```{r}
graph %>% 
  select_nodes("name == 'Louisa'") %>%
  trav_out_edge() %>%
  cache_edge_attrs_ws("commits", "numeric") %>%
  get_cache() %>% 
  sum
```

As a bit of an aside, we can use selections and rescale values to a styling attribute such as edge width, node size, or color. In this example, we can use the `rescale_edge_attrs()` function. It does a few things at once:

1. selects an entire column for an edge attribute in the graph's edge data frame
1. applies a rescaling of all the values in that column
1. optionally copies those values to a new edge attribute column, retaining the original values and edge attribute

Here, we will take the `commits` edge attribute and scale all values to the range `0.5` and `3.0` and use those values as the new `width` attribute (an aesthetic edge attribute that sets the width of each edge line).

```{r}
graph_scale_width_edges <-
  graph %>% 
  rescale_edge_attrs(
    "commits", 0.5, 3.0, "width")
```

It's almost always a good idea to check the node data frame or edge data frame after a transformation (as a sort of sanity check). Inspect the graph's edf using the `get_edge_df()` function.

```{r}
get_edge_df(graph_scale_width_edges)
```

Upon rendering the graph, the larger edges (and arrows) will indicate larger numbers of `commits`.

```{r}
render_graph(graph_scale_width_edges, output = "visNetwork")
```

The `rescale_edge_attrs()` function also works to scale across two colors. Here we take the `commits` edge attribute as before but scale from the color `gray95` (light gray) to the color `gray5` (dark gray). The color values (represented as hexadecimal color codes) are placed in the `color` edge attribute column. Here, we'll transform the graph and also display the edf to show the values in the new `color` column. 

```{r}
graph_scale_color_edges <-
  graph %>% 
  rescale_edge_attrs(
    "commits", "gray75", "gray10", "color")

get_edge_df(graph_scale_color_edges)
```

Upon rendering this graph, we can see very clearly that the darker edges represent higher numbers of commits.

```{r}
render_graph(graph_scale_color_edges, output = "visNetwork")
```

Getting back to traversals, we can perform interesting combinations of `select_`s so that specific information can be retrieved. Consider the case of getting the names of people that are above age `32`. Start by selecting all nodes that represent people and not repositories (so, `type == person`), and then make another selection where the `age` attribute has a value greater than `32`. On the second `select_nodes()` call, use `intersect` for the `set_op` argument. This indicates that any set of nodes found must intersect with the first previous set (this essentially applies the base **R** function `intersect()`). This guarantees that for any nodes where age is greater than `32`, such nodes are of `type == person`. Following that, cache the values for the `name` node attribute, then immediately collect those values in a vector with `get_cache()`.

```{r}
graph %>% 
  select_nodes("type == 'person'") %>%
  select_nodes("age > 32", "intersect") %>%
  cache_node_attrs_ws("name") %>%
  get_cache()
```

Here's another problem: getting the total number of commits from all people involved with the `supercalc` repository. We know that people contributing to projects have edges from people to projects (and not the other way around). Start with the project itself by selecting it with `select_nodes` and using a node attribute and value pair (we know that all project names are unique, so only one project will be selected). Then, use the `trav_in_edge()` function. This will move the selection from the node to all the edges that are pointing *in*. The last three steps are:

1. Caching all of the `commits` values (explicitly as numeric) for each of the edges in the selection
1. Getting those cached values (the output object is now a numeric vector)
1. Passing the vector to the `sum()` function to get the sum of commits

```{r}
graph %>% 
  select_nodes("project == 'supercalc'") %>%
  trav_in_edge() %>%
  cache_edge_attrs_ws("commits", "numeric") %>%
  get_cache() %>% 
  sum()
```

Say you wanted to get the name of the person that committed the most to the `supercalc` project. There are a few ways to do this. Below is a traversal-heavy solution:

```{r}
graph %>% 
  select_nodes("project == 'supercalc'") %>%
  trav_in_edge() %>%
  cache_edge_attrs_ws("commits", "numeric") %>%
  trav_in_node() %>%
  trav_in_edge(paste("commits ==", max(get_cache(.)))) %>%
  trav_out_node() %>%
  cache_node_attrs_ws("name") %>%
  get_cache()
```

Some explanation is required here. Since we always start with a selection, such selection began with the entity that known: the node with the value `supercalc` as a `project` node attribute (could have also used `label == supercalc` to make this selection). Then, a traversal was made to all the inbound edges. All of the edges' values for `commits` was then cached for later use. The `trav_in_node()` function was used to traverse back to the `supercalc` node (think: traversing into the node from adjacent edges). Again, we'll traverse back to the same edges, except, this time we'll set a condition that the selection can only occur for edges where `commits == max([commits_from_those_edges])`. In this case, that's a single edge that's selected. Since that edge is joined to the person that committed most to the `supercalc` project, we will traverse to that node using `trav_out_node()`. Since that node is now selected, we can use `cache_node_attrs_ws()` (cache node attributes with selection) and specify that the value for the `name` attribute should be cached. Right away, we `get_cache()` and the answer is revealed.

Because the graph has a specified structure, this pipeline of graph operations can be used to answer the same question about any repository. Simply create a function and provide the input graph and the project name to query.

```{r}

# Create a function that gets the person that committed
# the most to a project
get_highest_committer <- function(graph, project_name) {
  
  graph %>% 
    select_nodes(paste0("project == '", project_name, "'")) %>%
    trav_in_edge() %>%
    cache_edge_attrs_ws("commits", "numeric") %>%
    trav_in_node() %>%
    trav_in_edge(paste("commits ==", max(get_cache(.)))) %>%
    trav_out_node() %>%
    cache_node_attrs_ws("name") %>%
    get_cache()
}

# Call the function for each of the repositories to get
# the name of the person that committed most to each project
get_highest_committer(graph, "supercalc")
get_highest_committer(graph, "randomizer")
get_highest_committer(graph, "stringbuildeR")
```

## Part 3. Additional Graph Data Questions and Possible Solutions

There are so many questions one can ask of their dataset and the graph model provides flexibility in attaining solutions. Here are a few more problems and their solutions using graph traversals.

What is the email address of the individual that contributed the least to the `randomizer` project?

```{r}
graph %>% 
  select_nodes("project == 'randomizer'") %>%
  trav_in_edge() %>%
  cache_edge_attrs_ws("commits", "numeric") %>%
  trav_in_node() %>%
  trav_in_edge(paste("commits ==", min(get_cache(.)))) %>%
  trav_out_node() %>%
  cache_node_attrs_ws("email") %>%
  get_cache()
```

Get a list of all email addresses for the contributors (but not the maintainers) of the `randomizer` and `supercalc` projects. Consecutive `select_nodes()` calls can be thought of as 'and'-type selections (since `set_op == "union"` in the default case for all `select_...()` functions).

```{r}
graph %>% 
  select_nodes("project == 'randomizer'") %>%
  select_nodes("project == 'supercalc'") %>%
  trav_in_edge("rel == 'contributor'") %>%
  trav_out_node() %>%
  cache_node_attrs_ws("email", "character") %>%
  get_cache()
```

Which committer to the `randomizer` project has the highest number of followers? Recall that `follower_count` is a node attribute.

```{r}
graph %>% 
  select_nodes("project == 'randomizer'") %>%
  trav_in() %>%
  cache_node_attrs_ws(
    "follower_count", "numeric") %>%
  select_nodes("project == 'randomizer'") %>%
  trav_in(paste("follower_count ==", max(get_cache(.)))) %>%
  cache_node_attrs_ws("name") %>%
  get_cache()
```

Which people have committed to more than one project? (The solution contains no traversals but it's still interesting because it's predicated on one knowing the graph model.)

```{r}
graph %>%
  select_nodes_by_degree("outdeg > 1") %>%
  cache_node_attrs_ws("name") %>%
  get_cache()
```

## Conclusion

Graph traversals are undoubtedly useful operations. The nomenclature can be a point of confusion at first (I admit that their function names require some thought as to what kind of traversal will occur). However, given time and some practice, traversals will enable you to glean the appropriate information from a graph with the greatest of ease. As a graph becomes ever more complicated with the additions of nodes and edges, traversal functions (coupled with knowledge of a graph's underlying data model) will show their utility and worth.
